---
title: "Gun Control"
output: html_document
---


```{r include=FALSE}
#Load packages
library(tidyverse)
library(rvest)
library(httr)
library(plotly)
```



```{r echo=FALSE}
#Gun application and approval data
clean_fun = function(address, area){
  readxl::read_xlsx(address, sheet = "Data", range = area) %>% 
  rename(state = X__1) %>% 
  janitor::clean_names()
}


total_pop = clean_fun("./data/population-in-the-states-of-the-us-as-of-2017.xlsx", "B5:C56")

approved_lic = clean_fun("./data/number-of-federal-firearms-licensees-in-the-us-in-2017-by-state.xlsx", "B5:C57") %>% 
  filter(state != "Other Territories")

back_check = clean_fun("./data/nics-background-checks-done-by-us-firearms-licensees-2017-by-state.xlsx", "B5:C56")

gun_lic = inner_join(approved_lic, back_check, by = "state") %>% 
  inner_join(total_pop, by = "state") %>% 
  mutate(approval_rate = number_of_federal_firearms_licensees/number_of_residents_in_millions/1000000,
         application_rate = number_of_background_checks/number_of_residents_in_millions/1000000)
```




```{r include= FALSE}
#Mortality data

# Detailed Firearm Mortality Datatsets by Race, Age
firearm_mortality = read_csv("./data/cdc_firearm_mortality_data.csv", na = "Unreliable") %>% 
	janitor::clean_names() %>% 
	select(-ten_year_age_groups_code, -injury_mechanism_all_other_leading_causes_code, -race_code, death_cause =  injury_mechanism_all_other_leading_causes) %>% 
	mutate(ten_year_age_groups = factor(ten_year_age_groups, levels = c("1-4 years", "5-14 years", "15-24 years", "25-34 years", "35-44 years", "45-54 years", "55-64 years", "65-74 years", "75-84 years", "85+ years")))
# Summary Firearm Mortality Dataset by Year
firearm_mortality_summary = readxl::read_excel("./data/cdc_firearm_all_ages.xlsx") %>% 
	janitor::clean_names() %>% 
	select(-year_code, -injury_mechanism_all_other_leading_causes_code, death_cause= injury_mechanism_all_other_leading_causes)

# Created table with state name and state abbreviations crosswalk
st_crosswalk = tibble(state = state.name) %>%
   bind_cols(tibble(abb = state.abb)) %>% 
	 bind_rows(tibble(state = "District of Columbia", abb = "DC"))
# Joined abbreviation dataset with summary firearm mortality dataset
clean_firearm_mortality = left_join(firearm_mortality_summary, st_crosswalk, by = "state") %>% 
	rename(state_abb = abb) %>% 
  filter(year == max(year))
```


```{r echo=FALSE}
###Gun climate data

# website URL
url = "https://lawcenter.giffords.org/scorecard/#rankings"
gun_climate_url = read_html(url)
# extract table and clean data
gun_climate_data = 
	gun_climate_url %>% 
  html_nodes(css = "table") %>% 
  .[[1]] %>% 
  html_table(header = TRUE) %>% 
  as.tibble() %>% 
	janitor::clean_names() %>% 
	rename(
		law_strength = gun_law_strength_ranked, 
		grade_2017 = x2017grade, 
		death_rate_rank = gun_death_rate_ranked, 
		death_rate = gun_death_rate_per_100k) %>% 
	mutate(
    grade_2017 = factor(grade_2017, 
                      levels = c("A", "A-", "B+", "B", "C+", "C", "C-", "D", "D-", "F")), 
    grade_2017 = fct_collapse(grade_2017, 
    													A = c("A", "A-"), 
    													B = c("B+", "B"), 
    													C = c("C+", "C", "C-"), 
    													D = c("D", "D-")), 
    state_abb = state.abb)

```





### Gun Control Dataset Information

To quantify how strict are gun controls in different states, we are using:

+ [NICS](https://www.statista.com/statistics/249687/number-of-background-checks-done-by-the-nics-in-the-us-by-state/) background check data to approximate how many people tried to apply for gun licenses in each state; 
+  [Federal firarm licensee](https://www.statista.com/statistics/215670/number-of-federal-firearms-licensees-in-the-us-by-state/) data to approximate how many people actually got approved for gun licenses in each state.
+ [The total population](https://www.statista.com/statistics/183497/population-in-the-federal-states-of-the-us/) in different states data to standardize number of approved license and application numbers in differently populated states. 

Variable information:

+ The approval rate uses the proportion of licensees in the total population of the state to indicate the difficulties in getting apprroved for a gun license.
+ The application rate uses the proportion of background checks in the total population of the state to indicate people's willingness to apply for a gun license.
+ The crude rate is the mortality rate due to gun violence.
+ A numerical variable for law strength is used to indicate legislation strength in different states; the higher the law_strength value, the more lax the gun laws of a given state are.
+ The District of Columbia is excluded from the analysis and the gun control panel focuses on 2017. 

```{r echo=FALSE, include = FALSE}
gun_control = gun_climate_data %>% 
  inner_join(gun_lic, by = "state") %>% 
  inner_join(clean_firearm_mortality, by = "state")


skimr::skim(gun_control) %>%
            select(variable:stat,value) %>%
            filter((stat != "hist") , (stat != "top_counts"),(type %in% c("numeric","integer"))) %>% 
            spread(key = stat,value = value) %>% 
  knitr::kable(digits = 1) 
```

```{r echo=FALSE, include = FALSE}
skimr::skim(gun_control) %>%
            select(variable:stat,value) %>%
            filter((stat != "hist") , (stat != "top_counts"),(type %in% c("character","factor"))) %>% 
            spread(key = stat,value = value) %>% 
  knitr::kable(digits = 1) 

#As we can see, the dataset has no missing data. All the information, including population statistics, background check / applications statistics, approved licensees statistics,  are generalized for different states (except for D.C.) are inlcuded 
```

<br>

### Approval rate vs. Law strength

This plot shows the relationships between approval rate, mortality rate, and law strength in each state. 

<center> 

```{r echo=FALSE}

gun_control %>% 
  mutate(text_label = str_c("State:",state,  "\nCrude Rate:", crude_rate)) %>% 
  plot_ly(x = ~approval_rate, y = ~law_strength, 
          type = "scatter", mode = "markers", marker = list(size = ~crude_rate),
          alpha = 0.8, 
          color = ~law_strength,
          text = ~ text_label) %>%
  layout(
    title = "Approval percentage vs. Law Strength",
    xaxis = list(title = "Percentage of people approved for guns among total population"), 
    yaxis = list(title = "Law Strength"),
    annotations = list(
      x = 0.0012,
      y = 55,
      text = "Size of dots shows crude rate",
      xref = "x",
      yref = "y",
      ax = 0,
      ay = 0
))
         
```

</center> 

*Comments:*

Across different states, the approval rate for gun licenses is positively correlated with law strength. This is intuitive because we would expect it to be easier to be approved for a gun in states with less stringent gun laws. Also, with laxer gun laws states generally have higher mortality rates. A regression model was built to further explain this relationship, and information on this can be found through the [Regression Analysis](zanis_regression.html) tab.

<br>

## Application rate vs. Law strength

This plot shows the relationship between application rate, mortality rate, and law strength in each state.

<center> 

```{r echo=FALSE}

gun_control %>% 
  mutate(text_label = str_c("State:",state, '\nCrude Rate: ', crude_rate)) %>% 
  plot_ly(x = ~application_rate, y = ~law_strength,
          type = "scatter", mode = "markers", marker = list(size = ~crude_rate),
          alpha = 0.8, 
          color = ~law_strength,
          text = ~ text_label) %>%
  layout(
    title = "Application percentage vs. Law Strength",
    xaxis = list(title = "Percentage of people applied for guns among total population"), 
    yaxis = list(title = "Law Strength"),
    annotations = list(
  x = 0.8,
  y = 55,
  text = "Size of dots shows crude rate",
  xref = "x",
  yref = "y",
  ax = 0,
  ay = 0
))

```

</center> 

*Comments:*

+ The percentage of people who tried to apply for gun licensing is relatively consistent across states, regardless of law strength in a given state. This shows that **people across the country share an interest and enthusiasm in guns**.  
+ The notable exception is **Kentucky**. According to [Wikipedia](https://en.wikipedia.org/wiki/Gun_laws_in_Kentucky), people do not need a license or a permit to own guns for private use in Kentucky. This explains the exceptional passion to apply for guns in that locale, leading to an exceptionally high application rate. 

