---
title: "Crude Death Rate Prediction Using Economic, Legal and Health-Related Predictors"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# loading necessary packages
library(tidyverse)
library(PerformanceAnalytics)
library(modelr)
```

<br>

### Method

Data from several sources were joined together into a merged dataset. We use 2016 year to build the model. Main outcome is crude death rate for each state, candidate predictors are law strength, unemployment rate, sleep time, smoking, self-reported health, overweight, population, poverty, mental health, leisure physical activity, drinking, disability and diabetes for each state. Two models were generated using criteria-based model selection and stepwise regression. Models were compared by BIC, adjusted R square, Cp, etc. Models were examined by distribution of residuals (QQ plot, residuals vs fitted value), outliers. We also used cross-validation to compare the two models.

<br>

### Result

#### Correlation Matrix

<center>

```{r correlation_table, message = FALSE, warning = FALSE}
# read tidy dataset
data_for_reg <- read_csv("./data/merged_data.csv") %>%
	filter(year == 2016) %>% 
	spread(key = type_variable, value = Statistics) %>% 
	janitor::clean_names() %>% 
	filter(complete.cases(.)) %>% 
	# large number means low strength rank, so flip it
	mutate(law_strength_2016_only = 51 - law_strength_2016_only) %>% 
	select(-state, -year, -deaths)

# show correlation matrix
chart.Correlation(data_for_reg)
```

</center>

*Comments:*

Strong correlations can be seen among several pairs of variables. To minimize multi-collinearity, model should be selected carefully.

Most of the variates follow approximately normal distributions.

<br>

#### "Best" model for a given number of variables

```{r generate_models, results = FALSE}
# stepwise regression
crude_rate_reg <- lm(data = data_for_reg, formula = crude_rate ~ .)
stepwise_lm <- step(crude_rate_reg, direction = "backward")

# generate best subset for predictor number less than 8
(criteria_reg <- leaps::regsubsets(crude_rate ~ ., data = data_for_reg) %>% summary())

```

<center>

```{r criteria_based}
# generate "best subset" table
criteria_df <- as_tibble(criteria_reg$outmat) %>%
	mutate("n_pred" = row_number()) %>%
	select(n_pred, everything()) %>% 
	t()

colnames(criteria_df) <- c(1:8)
criteria_df[-1, ] %>% knitr::kable()
	
	
```

</center>

*Comments:*

It seems that "Law Strength", "Smoking", "Disability", and "Unemployment Rate" are all strong predictors. "Leisure & Physical Activities" appears in larger models.

<br>

#### R-squared, BIC, Cp, Regression Sum of Squares, Residual Sum of Squares

<center>

```{r model_selection}
# show model parameters for each model
tibble(
	n_pred = c(1:8),
	"Adj. R-sq" = criteria_reg$adjr2,
	"BIC" = criteria_reg$bic,
	"Cp" = criteria_reg$cp,
	"RSQ" = criteria_reg$rsq,
	"RSS" = criteria_reg$rss
	) %>%
	gather(key = "Statistics", value = "value", 2:6) %>%
	# plotting against number of predictors
	ggplot(aes(x = n_pred, y = value)) +
	  geom_point() +
	  geom_line() +
	  facet_grid(Statistics ~ ., scales = "free_y") +
	  labs(
	  	x = "Number of Predictors",
	  	y = "Values",
	  	title = "Model selection"
	  )
```

</center>

*Comments:*

The model with four predictors seems to have the highest adjusted R-squared value, lowest BIC and Cp values, and has a high regression mean sum of squares and low residual mean sum of squares. The six-predictor model has similar model statistics as the four-predictor model, with the exception of the higher BIC value. 

<br>

#### Two selected models of interest

**Model from stepwise selection:**

```{r stepwise}
# get statistics for model generated from stepwise regression
summary(stepwise_lm) %>% broom::tidy() %>% knitr::kable()
summary(stepwise_lm) %>% broom::glance() %>% knitr::kable()

```

*Comments:*

Drinking, law strength, and leisure all seem to be negatively associated with firearm crude death rate. Self-rated health, smoking, and unemployment rate are all positively associated with the firearm crude death rate.

**Criteria based model:**

```{r criteria}
# "best subset" according to model parameters
criteria_lm <- lm(data = data_for_reg, formula = crude_rate ~ disability +
										law_strength_2016_only + smoking + unemployment_rate)

# display statistics for criteria based "best subset" model
summary(criteria_lm) %>% broom::tidy() %>% knitr::kable()
summary(criteria_lm) %>% broom::glance() %>% knitr::kable()
```

*Comments:*

Disability, smoking, and unemployment rate are all  positively associated with the firearm crude death rate. Law strength is negatively associated with the firearm crude death rate.

<br>

#### Model diagnostics

<center>

*Criteria-based model (4 predictors)*

```{r model_diagnostics_citr}
# show residuals diagnosis
par(mfrow = c(2, 2))
plot(criteria_lm)
```

<br>

*Stepwise model (6 predictors)*

```{r model_diagnostics_stpws}
# show residuals diagnosis
par(mfrow = c(2, 2))
plot(stepwise_lm)

```

</center>

*Comments:*
For both models, no severe outliers were observed. Residuals seem to follow a normal distribution. Residuals don't hold constant over fitted values, but are still clustered around zero and not severely biased.

<br>

#### Cross validation

<center>

```{r crossvalidation}

set.seed(37)

# crossvalidation
crossv_mc(data_for_reg, 100) %>% 
	mutate(train = map(train, as_tibble),
				 test = map(test, as_tibble)) %>%
	mutate(stpws_mdl = map(train, ~lm(crude_rate ~ drinking +
																			law_strength_2016_only +
																			leisure_physical_activities +
																			self_reported_health +
																			smoking +
																			unemployment_rate,
																		data = .x)),
				 triv_mdl = map(train, ~lm(crude_rate ~ 1,
																		data = .x)),
				 crt_mdl = map(train, ~lm(crude_rate ~ disability +
				 												 	law_strength_2016_only +
				 												 	smoking +
				 												 	unemployment_rate,
				 												 data = .x))) %>% 
	mutate(rmse_stpws = map2_dbl(stpws_mdl, test, ~rmse(model = .x, data = .y)),
				 rmse_triv = map2_dbl(triv_mdl, test, ~rmse(model = .x, data = .y)),
				 rmse_crt   = map2_dbl(crt_mdl, test, ~rmse(model = .x, data = .y))) %>% 
	select("Stepwise(6)" = rmse_stpws,
	       "Trivival" = rmse_triv,
	       "Criteria(4)" = rmse_crt) %>%
	gather(key = model, value = rmse) %>%
	ggplot(aes(x = model, y = rmse)) +
	  geom_violin() +
    labs(
      x = "Models",
      y = "Root Mean Squared Errors",
      title = "Cross-validation"
    )
    


```

</center>

*Comments:*

Our four-predictor model seems to have a slightly lower root-mean-squared error (RMSE) value than the six-predictor model. They are both better than the trivial model of y ~ 1. For parsimony, we would ultimately choose the four-predictor model.

<br>

### Conclusion

Four-predictor model (disability, smoking, unemployment rate, law strength) seems to perform a little better than six-predictor model (drinking, law strength, leisure, self-rated health, smoking and unemployment rate). Model diagnosis shows that the residuals of both models agree with the underlying assumption. Cross validation shows a little higher rmse for six-predictor model than four-predictor model. Disability, smoking, unemployment rate positively associate with the firearm crude death rate. Law strength negatively associate with crude death rate.
